Convex optimisation in communications with cvxpy

### 2017-07-17 article_cvxpy.tex
cut down paper

\section{Real world applications}

\subsection{Maximising channel capacity}

Convex optimization can be used to find the channel capacity $C$ of a discrete memoryless channel. Consider a communication channel with input $X(t)\in\{1,2,...,n\}$ and output $Y(t)\in\{1,2,...m\}$. This means that the random variables $X$ and $Y$ can take $n$ and $m$ different values respectively.

In a discrete memoryless channel, the relation between the input and the output is given by the transition probability,
\begin{equation}
p_{ij} = \mathbb{P}(Y(t)=i|X(t)=j),
\end{equation} where these transition probabilities form the channel transition matrix $P$, with $P \in \mathbb{R}^{m\times n}$.  Assume that $X$ has a probability distribution denoted by $x\in\mathbb{R}^n$, meaning that,
\begin{equation}
x_j = \mathbb{P}(X(t) = j), \quad j=\{1,...,n\}.
\end{equation}
From Shannon \cite{shannon1949}, the channel capacity is given by the maximum possible mutual information, $I$, between $X$ and $Y$,
\begin{align}
C &= \sup_x I(X;Y), \quad \text{where},\\
I(X;Y) &= -\sum_{i=1}^{m}y_i\log_2 y_i+\sum_{j=1}^{n}\sum_{i=1}^{m}x_j p_{ij}\log_2 p_{ij}.
\end{align}

Given that $x\log x$ is convex for $x\geq 0$, we can formulate this as a convex optimisation problem,
\begin{align}
\text{minimise} \quad -I(X;Y), \\
\text{subject to} \quad \sum_{i=1}^{n}x_i = 1, \quad x \succeq 0.
\end{align} Here the constraints arise from $x$ describing a probability.

\subsubsection{Example \texttt{cvxpy} code}
Due to the built-in entropy function in \texttt{cvxpy}, \texttt{cvx.entr()}, the channel capacity problem is easy to write in DCP.

\lstinputlisting[language=Python]{channel_capacity.py}

\subsubsection{Computational complexity}
The computational complexity of the channel capacity problem follows a power law in \texttt{cvxpy}.  For $n$ up to 1000, the complexity is of order $\sim O(n^{2.5})$, Fig.~\ref{fig:channel_capacity_complexity}. Given that we are applying matrix multiplication to the power vector $\textbf{p}$, a complexity of at least $n^2$ would be expected.
\begin{figure}
\includegraphics[width=0.9\linewidth]{channel_capacity_complexity.pdf}
\caption{\label{fig:channel_capacity_complexity}Log-log plot of computation time (wall time) of channel capacity against problem size, for 1000 runs.}
\end{figure}

\subsection{Minimum power for a given SINR}
How can we minimise the total power consumption $P$ of $n$ transmitters each with power $p_i$, yet achieve a minimum SINR $\gamma_0$ for all $m$ receivers? This question is relevant to a telecoms company who want to offer a service at a minimum quality standard. We formulate the problem with a given square path gain matrix $G$, background noise level $\sigma$ and maximum power value that any transmitter can reach $P_{\text{max}}$:
		\begin{align*}
				&\underset{p}{\text{minimise}} \quad &&\sum_j p_j\\
				&\text{subject to} \quad &&p_j \leq P_{\max}\\
				& \quad &&p_j \geq 0\\
				& \quad &&\gamma_i \geq \gamma_0&&
		\end{align*}
(where symbols $\succ$, $\succeq$ denote elementwise inequalities between vectors or matrices.)

The desired signal to receiver $i$ is denoted by $S_i = G_{ik}p_k$, while the interference to $i$ is $I_i = \sum_{j\neq k}G_{ij}p_j$. However DCP does not allow division of the optimisation variable $p_i$ so  
\begin{equation*}
  \gamma_i \geq \gamma_0\Longleftrightarrow  \frac{S_i}{\sigma_i + I_i}\geq \gamma_0, \quad \forall i
\end{equation*}
was rearranged to:
\begin{equation*}
S_i-\gamma_0(\sigma_i + I_i)\geq 0, \quad \forall i
\end{equation*} which is DCP.
\subsubsection{Path gain}
The path gain $G_{ij}$ represents the proportion of power that reaches receiver $i$ from transmitter $j$. Supposing that the desired signal to receiver $i$ is from transmitter $k$, the power received is,
\begin{equation}
p_{ik}^{\text{rec}} = G_{ik}p_k.
\end{equation}
The signal-to-interference-plus-noise ratio (SINR) is the strength of the desired signal $S_i$ relative to the interference power $I_i$ plus the background noise $\sigma_i$ at $i$. Denoting the SINR at $i$ by $\gamma_i$, we have
\begin{equation}
\gamma_i = \frac{G_{ik}p_k}{\sigma_i+\sum_{j\neq k}G_{ij}p_j}
=\frac{S_i}{\sigma_i+I_i}.
\end{equation}
In a physical context, the path gain from transmitter $j$ to receiver $i$ will depend on the distance, $d_{ij}$ between them.  Assuming isotropic propagation from each transmitter, the fraction of power from $j$ that reaches $i$ is given by,

@@

\subsubsection{Example code}
To indicate the brevity of code implementation in \texttt{cvxpy} we show an abridged version below: \lstinputlisting[language=Python]{minimisePgivenSINR.py}

\subsubsection{Computational Complexity}

\begin{figure}[H]
\includegraphics[width=0.9\linewidth]{magazineMinSINR.pdf}
\caption{\label{fig:MinSINR}Log-log plot of computation time (wall time) of minimum power problem against the problem size, for 25 runs.}
\end{figure}

The computational complexity of the power minimisation for $n$ up to 1000 is of order $\sim O(n^{2.8})$.

\section{Conclusions}
This article has shown that many well-known convex problems in communications can be written in a DCP format, meaning that they can be easily implemented in \texttt{cvxpy}.

@@

The computational complexity of the problems examined had a clear trend which aligned with expectation.  This shows that convex optimisation solvers are robust to a variety of problem sizes and are scalable to at least problems of size 1000.

>>>>

\section{Minimum power for a given SINR}
How can we minimise the total power consumption $P$ of $n$ transmitters each with power $p_i$, yet achieve a minimum SINR $\gamma_0$ for all $m$ receivers? This question is relevant to a telecoms company who want to offer a service at a minimum quality standard. We formulate the problem with a given square path gain matrix $G$, background noise level $\sigma$ and maximum power value that any transmitter can reach $P_{\text{max}}$:
		\begin{align*}
				&\underset{p}{\text{minimise}} \quad &&\sum_j p_j\\
				&\text{subject to} \quad &&p_j \leq P_{\max}\\
				& \quad &&p_j \geq 0\\
				& \quad &&\gamma_i \geq \gamma_0&&
		\end{align*}
(where symbols $\succ$, $\succeq$ denote elementwise inequalities between vectors or matrices.)

The desired signal to receiver $i$ is denoted by $S_i = G_{ik}p_k$, while the interference to $i$ is $I_i = \sum_{j\neq k}G_{ij}p_j$. However DCP does not allow division of the optimisation variable $p_i$ so  
\begin{equation*}
  \gamma_i \geq \gamma_0\Longleftrightarrow  \frac{S_i}{\sigma_i + I_i}\geq \gamma_0, \quad \forall i
\end{equation*}
was rearranged to:
\begin{equation*}
S_i-\gamma_0(\sigma_i + I_i)\geq 0, \quad \forall i
\end{equation*} which is DCP.
\subsection{Path gain}
The path gain $G_{ij}$ represents the proportion of power that reaches receiver $i$ from transmitter $j$. Supposing that the desired signal to receiver $i$ is from transmitter $k$, the power received is,
\begin{equation}
p_{ik}^{\text{rec}} = G_{ik}p_k.
\end{equation}
The signal-to-interference-plus-noise ratio (SINR) is the strength of the desired signal $S_i$ relative to the interference power $I_i$ plus the background noise $\sigma_i$ at $i$. Denoting the SINR at $i$ by $\gamma_i$, we have
\begin{equation}
\gamma_i = \frac{G_{ik}p_k}{\sigma_i+\sum_{j\neq k}G_{ij}p_j}
=\frac{S_i}{\sigma_i+I_i}.
\end{equation}
In a physical context, the path gain from transmitter $j$ to receiver $i$ will depend on the distance, $d_{ij}$ between them.  Assuming isotropic propagation from each transmitter, the fraction of power from $j$ that reaches $i$ is given by,

@@

\subsection{Example code}
To indicate the brevity of code implementation in \texttt{cvxpy} we show an abridged version below: \lstinputlisting[language=Python]{minimisePgivenSINR.py}

\section{Conclusions}
This article has shown that a well-known convex problem in communications can be written in a DCP format, meaning that they can be easily implemented in \texttt{cvxpy}.

@@
